An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models

### 2025-05-02 Project_0/manuscript/manuscript.tex
saving work on project 0 edits
project 0 draft to submit

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Factors affecting reliablity of state-space age-structured assessment models},
  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Chengxue Li2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{url}
\usepackage{setspace}
%\singlespacing
%\onehalfspacing
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{relsize}
\usepackage{float}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{amsmath,rotating}
\usepackage[scanall]{psfrag}
\usepackage{bm}
\usepackage{caption,graphics}
\usepackage{graphicx}
\usepackage{sectsty}
\usepackage{color}
\usepackage{fancyhdr}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{upgreek}
\renewcommand\figurename{Fig.}
\captionsetup{labelsep=period, singlelinecheck=false}
\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
\renewcommand{\arraystretch}{1.5}
%\renewcommand\theadfont{}

@@

\title{Factors affecting reliablity of state-space age-structured
assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{08 April, 2025}

@@

\(^1\)corresponding author:
\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}\\
\(^2\)Northeast Fisheries Science Center, Woods Hole Laboratory, 166
Water Street, Woods Hole, MA 02543 USA\\
\(^3\)Woods Hole Oceanographic Institution\\
\(^4\)SMAST\\
\(^5\)Mid-Atlantic Fishery Management Council, 800 North State Street,
Suite 201, Dover, DE 19901 USA\\
\(^6\)Institute of Marine Research, Nye Flødevigveien 20, 4817 His,
Norway\\
\(^7\)Department of Ecology, Evolution, and Natural Resources. Rutgers
University\\

@@

Despite mounting efforts, several limitations remain. First, confounding
processes that can be treated as random effects in the model were not
thoroughly examined or tested within a simulation-estimation framework.
Second, previous studies relied on operating models conditioned on
specific fisheries, limiting their generalizability (Li et al. In
review; Liljestrand et al. 2024). In particular, the effects of
observation error and underlying fishing history have not been fully
isolated in simulation study designs, making it challenging to
disentangle the interplay between process and observation error
magnitudes, as demonstrated in Fisch et al. (2023). Third, explicitly
modeling stock-recruit relationships (SRRs) as mechanistic drivers of
population dynamics is promising (Fleischman et al. 2013; Du Pontavice
et al. 2022), but can reliability of inferences within integrated
state-space age-structured models has not been evaluated. Evidence from
other studies suggests that when both process and observation errors are
unknown, estimating density dependence parameters becomes highly
uncertain (Knape 2008; Polansky et al. 2009). In particular, Knape
(2008) demonstrated that stronger density dependence becomes
increasingly difficult to estimate in the presence of observation error.
Therefore, it is crucial to assess whether density dependence mechanisms
can be estimated with sufficient precision for use in fisheries
management (Auger-Méthé et al. 2016).

In the present study, we conduct a simulation study with operating
models (OMs) varying by degree of observation error, source and
variability of process error, and fishing history. The simulations from
these OMs are fitted with estimation models (EMs) that make alternative
assumptions for sources of process error, whether a SRR was estimated,
and whether a constant, or, in some EMs, median, natural mortality is
estimated. Given the confounding nature of process errors, developing
diagnostic tools to detect model misspecification is of great scientific
interest and could aid the next generation of stock assessments
(Auger-Méthé et al. 2021). We evaluate whether convergence and Akaike
Information Criterion (AIC) can correctly determine the source of
process error and the existence of a SRR. We also evaluate when
retrospective patterns occur and the degree of bias in the outputs of
the assessment model that are important for management.

@@

We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs
in our simulation study (Miller and Stock 2020; Stock and Miller 2021).
WHAM is an R package freely available via a github repository and is
built on the Template Model Builder package (Kristensen et al. 2016).
For this study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. WHAM has also been used to configure OMs
and EMs for closed loop simulations evaluating index-based assessment
methods (Legault et al. 2023) and is currently used or accepted for use
in management of numerous fish stocks in the Northeast United States
(NEUS) (e.g., NEFSC 2022a, 2022b; NEFSC 2024).

We completed a simulation study with a number of OMs that can be
categorized based on where process error random effects were assumed:
recruitment (R, assumed present in all models), apparent survival
(denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or
index catchability (R+q). We refer to the (R+S) OMs as modeling apparent
survival because on logscale the random effects (\(\epsilon_{a,y}\)) are
additive to the total mortality (F+M) between numbers at age, thus they
modify the survival term. However, as Stock and Miller (2021) note,
these random effects can be due to events other than mortality, such as
immigration, emigration, missreported catch, and other sources of
misspecification. For each OM, assumptions about the magnitude of the
variance of process errors and observations are required and the values
we used were based on a review of the range of estimates from NEUS
assessments using WHAM in stocks NEUS.

@@

We did not use the log-normal bias-correction feature for process errors
or observations described by (Stock and Miller 2021) for OMs and EMs to
simplify interpretation of the study results (Li et al. In review).
Simulations and model fitting were all carried out on the University of
Massachusetts Green High-Performance Computing Cluster. All code we used
to perform the simulation study and summarize results can be found at
\url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.

@@

We assumed a Beverton-Holt SRR with constant pre-recruit mortality
parameters for all OMs. All post-recruit productivity components are
constant in the apparent survival (R+S) and survey catchability (R+Sel)
process error OMs. Therefore, steepness and unfished recruitment are
also constant over the time period for those OMs (Miller and Brooks
2021). We specified unfished recruitment equal to \(e^{10}\) and
\(\Fmsy = F_{40\%} = 0.348\), which equates to a steepness of 0.69 and
\(a=0.60\) and \(b = 2.4 \times 10^{-5}\) for the Beverton-Holt
parameterization \[
N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
\] (Figure \ref{om_inputs_fig}, bottom right). For OMs without process
errors on natural mortality we fixed the rate at 0.2. For OMs with
process errors on natural mortality, the median natural mortality rate
was specified to be 0.2.

@@

All R+M OMs treat natural mortality as constant across age, but with
annually varying random effects. WHAM treats natural mortality as a
log-transformed parameter \[
\log M_{y,a} = \mu_{M} + \epsilon_{M,y}
\] that is a linear combination of a mean log-natural mortality
parameter that is constant across ages (\(\mu_{M} = \log(0.2)\)) and any
annual random effects are marginally distributed as
\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). Uncorrelated
random effects were also included on recruitment with \(\sigma_R = 0.5\)
(hence, we denote these OMs as R+M). The marginal standard deviations we
assumed for log natural mortality random effects were
\(\sigma_M \in \{0.1, 0.5\}\) and AR1 autocorrelation parameters of
\(\rho_M \in \{0,0.9\}\). The full factorial combination of these
process error assumptions and fishing history (2 levels) and observation
error (2 levels) scenarios described above results in 16 different R+M
OMs (Table \ref{M_om_table}).

@@

\hypertarget{convergence}{%
\subsubsection{Convergence}\label{convergence}}

@@

For a given model attribute we calculated the relative error \[
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
\] from fitting a given estimating model to simulated data set \(i\)
configured for a given OM where \(\widehat \theta_i\) and \(\theta_i\)
are the estimated and true values for simulation \(i\). We estimated
bias as the median of the relative errors across all simulations for a
given OM and EM combination. We constructed 95\% confidence intervals
for the median relative bias using the binomial distribution approach as
in Miller and Hyun (2018) and Stock and Miller (2021). We present
results for bias in terminal year estimates of SSB and recruitment,
Beverton-Holt stock recruit parameters (\(a\) and \(b\)), and median
natural mortality rate. Results for terminal year fishing mortality are
were strongly negatively correlated with those for SSB and are provided
in the Supplementary Materials.

@@

For R and R+S OMs, convergence generally declined for most EMs when the
median natural mortality rate was estimated and/or the Beverton-Holt SRR
was estimated even when the process error assumptions of the EMs and OMs
matched (Figure \ref{hessian_SE_convergence}, A). When there was high
observation error and constant fishing pressure (\(F = F_{\text{MSY}}\)
for all 40 years), convergence was poor for of all EM process error
configurations other than R EMs when fitted to R OMs
(\(\sigma_{2+} = 0\)) regardless of whether median natural mortality and
SRRs were estimated. Convergence of R EMs was high for all R and R+S OMs
except when there was high observation error and constant fishing
pressure, and when median natural mortality and SRRs were estimated. R+S
EMs fit to R OMs exhibited poor convergence regardless of whether
natural mortality or a SRR was estimated. R+S EMs fit to R+S OMs had
highest convergence rates when there was contrast in fishing pressure
and low observation error. Convergence rates were high for all EMs when
fit to data from R+S OMs with lower observation error except those where
median natural mortality and/or SRRs were estimated.

@@

Across all OMs, there was generally better accuracy (less bias and/or
lower variability) for estimation of the Beverton Holt \(a\) parameter
than the \(b\) parameter. In R and R+S OMs, EMs with the correct
assumptions about process errors provided the least biased estimation of
Beverton-Holt SRR parameters when there was a change in fishing pressure
over time and lower variability of recruitment process errors, but there
was little effect of estimating median natural mortality and a small
increase in bias for those OMs that had high observation error (Figure
\ref{SR_rel_error}, A). For other R and R+S OMs, estimating natural
mortality often resulted in less biased estimation of SRR parameters.
There was generally large variability in relative errors of the SRR
parameter estimates, but the lowest variability occurred with low
variability in recruitment and little or no variability in survival
process errors (\(\sigma_{2+} \in \{0,0.25\}\)), and contrast in fishing
pressure.

@@

Often poor convergence result when parameter estimates are at their
bounds (Carvalho et al. 2021), and this also applies to variance
parameters for random effects with state-space assessment models. Even
when Hessian is invertible, parameters that are poorly informed will
have extremely large variance estimates. This further inspection can
lead to a more appropriate and often more parsimonious model
configuration where the problematic parameters are not estimated. For
example, process error variance parameters that are estimated close to 0
indicates that the random effects are estimated to have little or no
variability and removing these process errors is warranted. Generally,
our results suggest we can expect lower probability of convergence of
state-space assessment models when estimating natural mortality or SRRs
because of the difficulty distinguishing these parameters from others
being estimated in assessment model with data that are typically
available. Our experiments did not aim to emulate the practitioner
decision process in developing model configurations (e.g.~removing a
source of process error and refitting the model when process error
variance parameters were estimated close to 0). Evaluating the efficacy
of such a decision process when applying Ems might be important in
closed loop simulations (e.g.~MSE) aimed at quantifying management
performance.

A factor affecting the convergence criteria, particularly for maximum
likelihood estimation of models with random effects, is numerical
accuracy. All optimizations performed in these simulations are of the
Laplace approximation of the marginal likelihood and, therefore,
gradients and Hessians are also with respect to this approximation (see
TMB::sdreport in the Template Model Builder package). Functionality
within the Template Model Builder package exists (i.e.,
TMB::checkConsistency) to check the validity of the Laplace
approximation and the utility of this as a diagnostic for state-space
assessment models should be explored further. Furthermore, numerical
methods are used to calculate and invert the Hessian for variance
estimation for models with random effects. Along with our results, the
potential lack of accuracy imposed by these approximations, suggests at
least investigating whether the Hessian is positive definite when the
calculated absolute gradients are not terribly large.

@@

Of the OM process error configurations we considered, we found AIC to be
accurate for selecting models with process errors on recruitment and
apparent survival (R and R+S). Fitting models to other OMs rarely
preferred R+S EMs, and R and R+S EMs were nearly always selected for the
matching OMs; a similar result was reported by Liljestrand et al.
(2024). For other sources of process error, accuracy of AIC was improved
to useful levels when there was larger variability in the process errors
and/or lower observation error.

@@

Although we did not compare models with alternative SRRs (e.g., Ricker
and Beverton-Holt), we do not expect AIC to perform any better
distinguishing between relationships. Our finding that AIC tended to
choose simpler recruitment models in most cases contrasts with the noted
bias in AIC for more complex models (Shibata 1976; Katz 1981; Kass and
Raftery 1995), but, whereas those findings apply to more the much more
common comparison of models that are fit to raw and independent
observations, here we are comparing state-space models which account for
observation error and estimate process errors in latent variables.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models},
  pdfauthor={Timothy J. Miller1,2; Greg Britten3; Elizabeth N. Brooks2; Gavin Fay4; Alex Hansell2; Christopher M. Legault2; Chengxue Li2; Brandon Muffley5; Brian C. Stock6; John Wiedenmann7},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{url}
\usepackage{setspace}
%\singlespacing
%\onehalfspacing
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{relsize}
\usepackage{float}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{amsmath,rotating}
\usepackage[scanall]{psfrag}
\usepackage{bm}
\usepackage{caption,graphics}
\usepackage{graphicx}
\usepackage{sectsty}
\usepackage{color}
\usepackage{fancyhdr}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{upgreek}
\renewcommand\figurename{Fig.}
\captionsetup{labelsep=period, singlelinecheck=false}
\newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
\renewcommand{\arraystretch}{1.5}
%\renewcommand\theadfont{}

@@

\title{An investigation of factors affecting inferences from and
reliability of state-space age-structured assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{02 May, 2025}

@@

\(^1\)corresponding author:
\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}\\
\(^2\)Northeast Fisheries Science Center, Woods Hole Laboratory, 166
Water Street, Woods Hole, MA 02543 USA\\
\(^3\)Biology Department, Woods Hole Oceanographic Institution, 266
Woods Hole Rd. Woods Hole, MA, USA\\
\(^4\)Department of Fisheries Oceanography, School for Marine Science
and Technology, University of Massachusetts Dartmouth, 836 S Rodney
French Boulevard, New Bedford, MA 02740, USA\\
\(^5\)Mid-Atlantic Fishery Management Council, 800 North State Street,
Suite 201, Dover, DE 19901 USA\\
\(^6\)Institute of Marine Research, Nye Flødevigveien 20, 4817 His,
Norway\\
\(^7\)Department of Ecology, Evolution, and Natural Resources. Rutgers
University\\

@@

Despite mounting efforts, several limitations remain. First, confounding
processes that can be treated as random effects in the model were not
thoroughly examined or tested within a simulation-estimation framework.
Second, previous studies relied on operating models conditioned on
specific fisheries, limiting their generalizability (Li et al. In
review; Liljestrand et al. 2024). In particular, the effects of
observation error and underlying fishing history have not been fully
isolated in simulation study designs, making it challenging to
disentangle the interplay between process and observation error
magnitudes, as demonstrated in Fisch et al. (2023). Third, explicitly
modeling stock-recruit relationships (SRRs) as mechanistic drivers of
population dynamics is promising (Fleischman et al. 2013; Pontavice et
al. 2022), but reliability of inferences within integrated state-space
age-structured models has not been evaluated. Evidence from other
studies suggests that when both process and observation errors are
unknown, estimating density dependence parameters becomes highly
uncertain (Knape 2008; Polansky et al. 2009). In particular, Knape
(2008) demonstrated that stronger density dependence becomes
increasingly difficult to estimate in the presence of observation error.
Therefore, it is crucial to assess whether density dependence mechanisms
can be estimated with sufficient precision for use in fisheries
management (Auger-Méthé et al. 2016). Finally, although the importance
of autocorrelation in process errors is recognized, investigations of
the ability to distinguish state-space assessment models with and
without autocorrelation and whether such misspecification is detrimental
to estimation of important population metrics are lacking (Johnson et
al. 2016; Xu et al. 2019).

In the present study, we conduct a simulation study with operating
models (OMs) varying by degree of observation error, source and
variability of process error, and fishing history. The simulations from
these OMs are fitted with estimation models (EMs) that make alternative
assumptions for sources of process error, whether a SRR was estimated,
and whether natural mortality is estimated. Given the confounding nature
of process errors, developing diagnostic tools to detect model
misspecification is of great scientific interest and could aid the next
generation of stock assessments (Auger-Méthé et al. 2021). We evaluate
whether convergence and Akaike Information Criterion (AIC) can correctly
determine the source of process error and the existence of a SRR. We
also evaluate when retrospective patterns occur and the degree of bias
in the outputs of the assessment model that are important for
management.

@@

We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs
in our simulation study (Miller and Stock 2020; Stock and Miller 2021).
WHAM is an R package freely available via a github repository and is
built on the Template Model Builder package (Kristensen et al. 2016).
For this study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. WHAM has also been used to configure OMs
and EMs for closed loop simulations evaluating index-based assessment
methods (Legault et al. 2023) and is currently used or accepted for use
in management of numerous NEUS fish stocks (e.g., NEFSC 2022a, 2022b;
NEFSC 2024).

We completed a simulation study with a number of OMs that can be
categorized based on where process error random effects were assumed:
recruitment (R, assumed present in all models), apparent survival
(denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or
index catchability (R+q). We refer to the (R+S) OMs as modeling apparent
survival because on logscale the random effects (\(\epsilon_{a,y}\)) are
additive to the total mortality (F+M) between numbers at age, thus they
modify the survival term. However, as Stock and Miller (2021) note,
these random effects can be due to events other than mortality, such as
immigration, emigration, missreported catch, and other sources of
misspecification. For each OM, assumptions about the magnitude of the
variance of process errors and observations are required and the values
we used were based on a review of the range of estimates from Northeast
Unite States (NEUS) assessments using WHAM.

@@

We did not use the log-normal bias-correction feature for process errors
or observations described by (Stock and Miller 2021) for OMs and EMs to
simplify interpretation of the study results (Li et al. In review). All
code we used to perform the simulation study and summarize results can
be found at
\url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.

@@

We assumed a Beverton-Holt SRR with constant pre-recruit mortality
parameters for all OMs. All biological inputs to calculations of
spawning biomass per recruit (i.e., weight, maturity, and natural
mortality at age) are constant in the apparent survival (R+S)
selectivity (R+Sel), and survey catchability (R+q) process error OMs.
Therefore, steepness and unfished recruitment are also constant over the
time period for those OMs (Miller and Brooks 2021). We specified
unfished recruitment equal to \(e^{10}\) and
\(\Fmsy = F_{40\%} = 0.348\), which equates to a steepness of 0.69 and
\(a=0.60\) and \(b = 2.4 \times 10^{-5}\) for the Beverton-Holt
parameterization \[
N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
\] (Figure \ref{om_inputs_fig}, bottom right). We assumed a value of 0.2
for the natural mortality rate in OMs without process errors on natural
mortaliyt and for the median rate for OMs with process errors on natural
mortality.

@@

All R+M OMs treat natural mortality as constant across age, but with
annually varying random effects. WHAM treats natural mortality as a
log-transformed parameter \[
\log M_{y,a} = \mu_{M} + \epsilon_{M,y}
\] that is a linear combination of a mean log-natural mortality
parameter that is constant across ages (\(\mu_{M} = \log(0.2)\)) and any
annual random effects are marginally distributed as
\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). The marginal
standard deviations we assumed for log natural mortality random effects
were \(\sigma_M \in \{0.1, 0.5\}\) and the random effects were either
uncorrelated or first-order autoregressive (AR1,
\(\rho_M \in \{0,0.9\}\)). Uncorrelated random effects were also
included on recruitment with \(\sigma_R = 0.5\) (hence, we denote these
OMs as R+M). The full factorial combination of these process error
assumptions and fishing history (2 levels) and observation error (2
levels) scenarios described above results in 16 different R+M OMs (Table
\ref{M_om_table}).

@@

\hypertarget{convergence}{%
\subsubsection*{Convergence}\label{convergence}}
\addcontentsline{toc}{subsubsection}{Convergence}

@@

For a given model attribute we calculated the relative error \[
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
\] from fitting a given estimating model to simulated data set \(i\)
configured for a given OM where \(\widehat \theta_i\) and \(\theta_i\)
are the estimated and true values for simulation \(i\). We estimated
bias as the median of the relative errors across all simulations for a
given OM and EM combination. We constructed 95\% confidence intervals
for the median relative bias using the binomial distribution approach as
in Miller and Hyun (2018) and Stock and Miller (2021). We present
results for bias in terminal year estimates of SSB and recruitment,
Beverton-Holt stock recruit parameters (\(a\) and \(b\)), and median
natural mortality rate. Results for terminal year fishing mortality were
strongly negatively correlated with those for SSB and are provided in
the Supplementary Materials.

@@

For many R and R+S OMs, convergence rate declined when either the median
natural mortality rate or the Beverton-Holt SRR was estimated even when
the process error assumptions of the EMs and OMs matched (Figure
\ref{hessian_SE_convergence}, A). When there was high observation error
and constant fishing pressure (\(F = F_{\text{MSY}}\) for all 40 years),
convergence was poor for of all EM process error configurations other
than R EMs when fitted to R OMs (\(\sigma_{2+} = 0\)) regardless of
whether median natural mortality and SRRs were estimated. Convergence of
R EMs was high for all R and R+S OMs except when there was high
observation error and constant fishing pressure, and when median natural
mortality and SRRs were estimated. R+S EMs fit to R OMs exhibited poor
convergence regardless of whether natural mortality or a SRR was
estimated. R+S EMs fit to R+S OMs had highest convergence rates when
there was contrast in fishing pressure and low observation error.
Convergence rates were high for all EMs when fit to data from R+S OMs
with lower observation error except those where median natural mortality
and/or SRRs were estimated.

@@

Across all OMs, there was generally less bias and(or) lower variability
in estimation of the Beverton Holt \(a\) parameter than the \(b\)
parameter. In R and R+S OMs, EMs with the correct assumptions about
process errors provided the least biased estimation of Beverton-Holt SRR
parameters when there was a change in fishing pressure over time and
lower variability of recruitment process errors, but there was little
effect of estimating median natural mortality and a small increase in
bias for those OMs that had high observation error (Figure
\ref{SR_rel_error}, A). For other R and R+S OMs, estimating natural
mortality often resulted in less biased estimation of SRR parameters.
There was generally large variability in relative errors of the SRR
parameter estimates, but the lowest variability occurred with low
variability in recruitment and little or no variability in survival
process errors (\(\sigma_{2+} \in \{0,0.25\}\)), and contrast in fishing
pressure.

@@

Often poor convergence result when parameter estimates are at their
bounds (Carvalho et al. 2021), and this also applies to variance
parameters for random effects with state-space assessment models. Even
when the Hessian is invertible, parameters that are poorly informed will
have extremely large variance estimates. This further inspection can
lead to a more appropriate and often more parsimonious model
configuration where the problematic parameters are not estimated. For
example, process error variance parameters that are estimated close to 0
indicates that the random effects are estimated to have little or no
variability and removing these process errors is warranted. Generally,
our results suggest we can expect lower probability of convergence of
state-space assessment models when estimating natural mortality or SRRs
because of the difficulty distinguishing these parameters from others
being estimated in assessment model with data that are typically
available. Our experiments did not aim to emulate the practitioner
decision process in developing model configurations (e.g.~removing a
source of process error and refitting the model when process error
variance parameters were estimated close to 0). Evaluating the efficacy
of such a decision process when applying EMs might be important in
closed loop simulations (e.g.~MSE) aimed at quantifying management
performance.

A factor affecting the convergence criteria, particularly for maximum
likelihood estimation of models with random effects, is numerical
accuracy. All optimizations performed in these simulations are of the
Laplace approximation of the marginal likelihood and, therefore,
gradients and Hessians are also with respect to this approximation (see
TMB::sdreport in the Template Model Builder package). Functionality
within the Template Model Builder package exists (i.e.,
TMB::checkConsistency) to check the validity of the Laplace
approximation and the utility of this as a diagnostic for state-space
assessment models should be explored further. Furthermore, numerical
methods are used to calculate and invert the Hessian for variance
estimation for models with random effects. Along with our results, the
potential lack of accuracy imposed by these approximations, suggests at
least investigating whether the Hessian is positive definite when the
calculated absolute gradients are not terribly large (e.g, \(< 1\)).

@@

Of the OM process error configurations we considered, we found AIC to be
accurate for selecting models with process errors on recruitment and
apparent survival (R and R+S). Fitting models to other OMs rarely
preferred R+S EMs, and R and R+S EMs were nearly always selected for the
matching OMs; a similar result was reported by Liljestrand et al.
(2024). For other sources of process error, accuracy of AIC was improved
when there was larger variability in the process errors and/or lower
observation error.

@@

Although we did not compare models with alternative SRRs (e.g., Ricker
and Beverton-Holt), we do not expect AIC to perform any better
distinguishing between relationships. Our finding that AIC tended to
choose simpler recruitment models in most cases contrasts with the noted
bias in AIC for more complex models (Shibata 1976; Katz 1981; Kass and
Raftery 1995), but, whereas those findings apply to the much more common
comparison of models that are fit to raw and independent observations,
here we are comparing state-space models which account for observation
error and estimate process errors in latent variables.

Our results comport with those of de Valpine and Hastings (2002) who
found AIC could not distinguish among state-space SRRs that were fit
just to SSB and recruitment observations (i.e., not an assessment
model). Similarly, Britten et al. (In review) found AIC could not
reliably distinguish alternative environmental effects on SRR
parameters. However, Miller et al. (2016) did find AIC to prefer a SRR
with environmental effects when applied to data for the SNEMA yellowtail
flounder stock and AIC also selected an environmental covariate on a SRR
for the most recent stock assessment of Georges Bank yellowtail flounder
(NEFSC 2025). Both of these yellowtail flounder stocks have large
changes in stock size and the values of environmental covariates over
time. Additionally, this species is well-observed by the bottom trawl
survey that is used for an index in assessment models.
